Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-06-03 article.tex
Possible extensions
Warn about reprocical
Discuss partial correlation in term of independance and relate to the literature

\paragraph{Problem definition\\}

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}

% introduce the notion of dependence / independence
From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and
independences.  Even if \textit{independence
does no imply causation}[todo cite], it gives insight on the underlying connectivity
network. Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
to a set of neurons $X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$, denoted
by  $X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,  if
\begin{align*}
&p_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i, x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) \\
=&p_{X_i | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
p_{X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
\end{align*}
for all real numbers $x_1,\ldots,x_p $ such that
$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
 (x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$.

@@

% Correlation => Issue indirect effect
% Partial correlation => better estimate of the graph  => Symmetric measure

Thus if $X_i \not\perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,
then we have the edge $i-j$ in the undirected graph.

@@

% Introduce partial correlation
\paragraph{Partial correlation\\}

A common association measure is the
\begin{align}
\rho_{ij} = \dfrac{cov(I,J)}{\sigma_I \sigma_J}
\end{align}
Even if \textit{correlation does not imply causation} [todo cite], it may give
an insight on the underlying connectivity network. Nevertheless, the main
challenge in inferring a network is to distinguish direct from indirect
interactions [de la Fuente et al., 2004]. While the classical correlation
coefficient characterize both direct and indirect interactions, the partial
correlation coefficient quantifies association between variables conditionally
on one or several other variables (the number of variables considered being
the order of the partial correlation coefficient). [Shipley, 2002] suggests to
consider correlation coefficients of all possible orders, i.e. conditioned
on every subsets of the set of $p-2$ other variables, in order to take into
account multiple indirect paths between a given pair of variables.  An obvious
limitation to calculate high-order coefficient is the number of samples in
conditioned data sets. Considering the large number of samples at our disposal
in the connectomics challenge, we only consider the highest order of partial
correlation coefficient. We thus try to estimate partial correlation matrix as
the inferred network where a matrix element, denoted by
$\rho_{ij | \mathcal{V}^{-i,j}}$, is the partial correlation coefficient of
order $p-2$ between neurons $i$ and $j$. Note that this matrix is symmetric
and thus it is impossible to determine the causal mechanism behind the
interaction (I directly causes J, J directly causes I or both
[de la Fuente, 2004])\footnote{Two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) [de la Fuente, 2004].} .
A way to find these coefficient [todo cite?] is to take the inverse of the
covariance matrix to get the partial correlation matrix of order $p-2$, also
known as the precision matrix [todo cite] .

% Introduce directivity
\paragraph{Directivity}

% Introduce filtering
\paragraph{Filtering\\}

@@

>>>>

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions [de la Fuente et al., 2004].
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

% introduce the notion of dependence / independence
From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and
independences.  Even if \textit{independence
does no imply causation}[todo cite], it gives insight on the underlying connectivity
network. For instance, the graph $A \rightarrow B \rightarrow C$ translates the
following dependances and independances: $A \not\perp B$, $A \not\perp B | C$,
$B \not\perp C$, $B \not\perp C | A$, but $A \perp C | B$ and  $A \not\perp C$.

Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
to a set of neurons $X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$, denoted
by  $X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,  if
\begin{align*}
&p_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i, x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) \\
=&p_{X_i | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
p_{X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
\end{align*}
for all real numbers $x_1,\ldots,x_p $ such that
$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
(x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$. Note that this
reciprocal is true only under additional hypotheses.

@@

% Correlation => Issue indirect effect
% Partial correlation => better estimate of the graph  => Symmetric measure
While the correlation coefficient is unable to distinguish between
direct and indirect interaction, the partial correlation coefficient
aims to infer only direct ones through the conditioning over all the other
variables. In previous studies [Shipley, 2002], it has been suggested
consider correlation coefficients of all possible orders [Shipley, 2002],
i.e. conditioned on every subsets of the set of $p-2$ other variables, in order
to take into account multiple indirect paths between a given pair of variables.
An obvious limitation to calculate high-order coefficient is the number of
samples in conditioned data sets \textcolor{red}{AJ: Est ce que ce ne sera pas
à la fois $p$ and $n$?}. Given the high number of samples $n$
in the Connectomics challenge, we only consider the highest order of partial
correlation coefficient.

Note that this matrix is symmetric
and thus it is impossible to determine the causal mechanism behind the
interaction (I directly causes J, J directly causes I or both
[de la Fuente, 2004])\footnote{Two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) [de la Fuente, 2004].}

@@

% Introduce directivity

% Introduce filtering

@@

% Possible extensions
% partial autocorrelation function (sometimes "partial correlation function")
% conditional independence test
